---
title: "ps05-code-czahn2"
author: "camille zahn"
date: "2023-02-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library("readr")
library("ggplot2")
```

```{r}
getwd()
```
***1 Load and check data (5pt)***

2. (2pt) Load data. How many rows/columns do we have?
```{r, message=FALSE}
library(tidyverse)
gapminder <- read_delim("gapminder.csv.bz2")
```
The _gapminder_ data contains `r nrow(gapminder)` rows and `r ncol(gapminder)` columns.

3. (2pt) Print a small sample of data. Does it look OK?
```{r}
gapminder %>%
  sample_n(10) %>%
  knitr::kable()
## Yes, looks good!
```
***2 Descriptive statistics (15pt)***

1. (3pt) How many countries are there in the dataset? Analyze all three: iso3, iso2 and name.
```{r}
num_iso2 <- gapminder %>%
  select(iso2) %>%
  summarize(n_distinct(iso2))
```
There are `r num_iso2` distinct iso2 codes
```{r}
num_iso3 <- gapminder %>%
  select(iso3) %>%
  summarize(n_distinct(iso3))
```
There are `r num_iso3` distinct iso3 codes
```{r}
num_name <- gapminder %>%
  select(name) %>%
  summarize(n_distinct(name))
```
There are `r num_name` distinct country names

2. If you did this correctly, you saw that there are more names than iso-2 codes, and there are
even more iso3 -codes. What is going on? Can you find it out?
(a) (5pt) Find how many names are there for each iso-2 code. Are there any iso-2 codes that
correspond to more than one name? What are these countries?
```{r}
gapminder %>%
  filter(!is.na(name)) %>%
  group_by(iso2) %>%
  summarize(name_count = n_distinct(name)) %>%
  arrange(desc(name_count))

gapminder %>%
  filter(!is.na(iso2)) %>%
  distinct(name)

```
(b) (5pt) Now repeat the same for name and iso3-code. Are there country names that have
more than one iso3-code? What are these countries?
Hint: two of these entitites are CHANISL and NLD CURACAO.
```{r}
gapminder %>%
  group_by(name) %>%
  select(iso3, name) %>%
  summarize(name_count = n_distinct(iso3)) %>%
  filter(name_count > 1)
```
There are 4 country names that have more than one iso-3 code.


3. (2pt) What is the minimum and maximum year in these data?
```{r}
max_time <- gapminder %>%
  select(time) %>%
  filter(!is.na(time)) %>%
  summarize(max(time))
```
The maximum year is `r max_time`
```{r}
min_time <- gapminder %>%
  select(time) %>%
  filter(!is.na(time)) %>%
  summarize(min(time))
```
The minimum year is `r min_time`
```{r}
```
***3 CO2 emissions (30pt)***

1. (2pt) How many missing co2 emissions are there for each year? Analyze both missing CO2
and co2_PC. Which years have most missing data?
```{r}
gapminder %>%
  group_by(time) %>%
  summarize(missing_count = sum(is.na(co2) | is.na(co2_PC))) %>%
  arrange(desc(missing_count)) %>%
  head()
```

2. (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few
more countries of your choice. Explain what do you see.

```{r}
gapminder %>%
  filter(iso3 %in% c("USA", "CHN", "IND", "RUS", "JPN")) %>%
  ggplot(aes(x = time, y = co2, color = name)) +
  geom_line() +
  labs(x = "Year", y = "CO2 Emissions (metric tons per capita)", color = "Country")
## all countries except Russia have increased emissions over time, and China has the most
```

3. (5pt) Now let’s analyze the CO2 emissions per capita (co2_PC ). Make a similar plot of the
same countries. What does this figure suggest?
```{r}
gapminder %>%
  filter(iso3 %in% c("USA", "CHN", "IND", "RUS", "JPN")) %>%
  ggplot(aes(x = time, y = co2_PC, color = name)) +
  geom_line() +
  labs(x = "Year", y = "CO2 Emissions (metric tons per capita)", color = "Country")

## Since China has such a large population, their per capita emissions is lower than that of 
## Japan, Russia, and the United States.
```

4. (6pt) Compute average CO2 emissions per capita across the continents (assume region is the
same as continent). Comment what do you see.
Note: just compute averages over countries and ignore the fact that countries are of different
size.
Hint: Americas 2016 should be 4.80.
```{r}
gapminder %>%
  group_by(region, time) %>%
  filter(!is.na(co2_PC), !is.na(name)) %>%
  filter(time == 1960 | time == 2016) %>%
  summarize(mean = mean(co2_PC))
## All regions except Americas increased co2 emissions from 1960-2019
```
5. (7pt) Make a barplot where you show the previous results–average CO2 emissions per capita
across continents in 1960 and 2016.
```{r}
avg_1960_2016 <- gapminder %>%
  group_by(region, time) %>%
  filter(!is.na(co2_PC), !is.na(name)) %>%
  summarize(mean = mean(co2_PC)) %>%
  filter(time == 1960 | time == 2016)
  ggplot(data = avg_1960_2016,
           aes(x = region, y = mean, fill = as.factor(time))) +
    geom_bar(stat = "identity", position = "dodge")

```

6. Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per
capita) in 2016 for each continent? (Assume region is continent).
```{r}
gapminder %>%
  select(time, region, co2_PC, name) %>%
  filter(time == 2016) %>%
  group_by(region, time) %>%
  summarize(top3 = paste(head(name[order(co2_PC, decreasing = TRUE)], 3), collapse = ", "),
            bottom3 = paste(head(name[order(co2_PC)], 3), collapse = ", "))
```
***4 GDP per capita (50pt)***

1. (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, using data for
1960. Make the point size dependent on the country size, and color those according to the
continent. Feel free to adjust the plot in other ways to make it better.
Comment what do you see there.
```{r}
gapminder_1960 <- filter(gapminder, time == 1960)

ggplot(gapminder_1960, aes(x = GDP_PC, y = lifeExpectancy, size = totalPopulation, col = region)) +
  geom_point() +
  scale_color_manual(values = c("red", "blue", "green", "purple", "orange", "brown")) +
  labs(title = "GDP per capita vs. Life expectancy by country in 1960", x = "GDP per capita", y = "Life expectancy")
## As GDP increases, life expectancy also tends to increase. The countries with small GPD and LE are mostly in Africa or Asia,
## and the countries with large GDP and LE are mostly in the Americas or Europe.
```
2. (4pt) Make a similar plot, but this time use 2019 data only.
```{r}
gapminder_2019 <- filter(gapminder, time == 2019)

ggplot(gapminder_2019, aes(x = GDP_PC, y = lifeExpectancy, size = totalPopulation, col = region)) +
  geom_point() +
  scale_color_manual(values = c("red", "blue", "green", "purple", "orange", "brown")) +
  labs(title = "GDP per capita vs. Life expectancy by country in 2019", x = "GDP per capita", y = "Life expectancy")
```
3. (6pt) Compare these two plots and comment what do you see. How has world developed
through the last 60 years?
```{r}
## The biggest difference i see in the plots is that many more Asian countries have higher GDP and LE in 2019.
## Every continent has increased, but Asia stands out to me the most.
```

4. (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results
fit with what do you see on the figures?
Note: here as average I mean just average over countries, ignore the fact that countries are of
different size.
```{r}
gapminder_1960 %>%
  group_by(region) %>%
  filter(!is.na(lifeExpectancy)) %>%
  summarize(mean = mean(lifeExpectancy))

gapminder_2019 %>%
  group_by(region) %>%
  filter(!is.na(lifeExpectancy)) %>%
  summarize(mean = mean(lifeExpectancy))
##Yes, average LE increased significantly for every continent between 1960-2019.
```

5. (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results
in the order of growth. Explain what do you see.
Hint: these data (data in long form) is not the simplest to compute growth. But you may
want to check out the lag() function. And do not forget to group data by continent when
using lag(), otherwise your results will be messed up! See https://faculty.washington.
edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute.
```{r}
gapminder %>%
  group_by(region) %>%
  arrange(name, time) %>%
  mutate(avg_le = (lifeExpectancy - lag(lifeExpectancy)) / lag(lifeExpectancy)) %>%
  filter(time >= 1961) %>%
  summarize(avg_growth = mean(avg_le, na.rm = TRUE)) %>%
  arrange(avg_growth)
```
6. (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both
histograms on the same graph, see how well you can do it!
```{r}
ggplot() + 
 geom_histogram(data = gapminder_1960, aes(x = GDP_PC), 
                 binwidth = 5000, fill = "blue", alpha = 0.5) +
geom_histogram(data = gapminder_2019, aes(x = GDP_PC), 
                 binwidth = 5000, fill = "red", alpha = 0.5) +
  xlab("GDP per capita") +
  ylab("Count") +
  ggtitle("Histogram of GDP per capita for 1960 and 2019") +
  scale_fill_manual(values = c("blue", "red"), 
                    labels = c("1960", "2019"), 
                    name = "Year") +
  # Set axis limits to zoom in on the data
  xlim(0, 50000) +
  ylim(0, 100)
```
7. (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When
counting from top.)
Hint: check out the function rank()!
Hint2: 17 for 1960.
```{r}
gapminder_ranked <- gapminder %>%
  filter(!is.na(time), !is.na(name)) %>%
  group_by(time, name) %>%
  mutate(rank = rank(desc(lifeExpectancy)))

us_rank_1960 <- gapminder_ranked %>%
  filter(time == 1960 & name == "United States") %>%
  pull(rank)

us_rank_2019 <- gapminder_ranked %>%
  filter(time == 2019 & name == "United States") %>%
  pull(rank)

# Print results
cat("Ranking of US in 1960:", us_rank_1960, "\n")
cat("Ranking of US in 2019:", us_rank_2019, "\n")
```
Comment: I couldn't get this one to function correctly :( Dont know why it won't print the ranking.

8. (6pt) If you did this correctly, then you noticed that US ranking has been falling quite a
bit. But we also have more countries in 2019–what about the relative rank divided by the
corresponding number of countries that have LE data in the corresponding year?
Hint: 0.0904 for 1960.
```{r}

# calculate relative rank for 1960
gapminder_1960 %>%
  mutate(rank = rank(desc(lifeExpectancy))) %>%
  filter(name == "USA")
rank_1960 <- gapminder_1960$rank / n_distinct(gapminder$name[gapminder$time == 1960])

# calculate relative rank for 2019
gapminder_2019 %>%
  mutate(rank = rank(desc(lifeExpectancy))) %>%
  filter(name == "USA")
rank_2019 <- gapminder_2019$rank / n_distinct(gapminder$name[gapminder$time == 2019])

# print results
cat("Relative rank of the US in terms of life expectancy in 1960:",
    round(rank_1960, 2), "\n")
cat("Relative rank of the US in terms of life expectancy in 2019:",
    round(rank_2019, 2), "\n")

```
Comment: I couldn't get this one to function correctly :( Dont know why it won't print the ranking.

I spent 10 hours on this PS.
